---
title: "YM-1C"
author: "Dylan Stark"
date: "2/20/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(tidytext)
library(modelr)

library(courseraswiftkey)
```

```{r}
data("en_us_news")
data("en_us_twitter")
data("en_us_blogs")
```

```{r}
add_boundaries <- function(data, start = "_s_start", stop = "_s_end") {
  data %>%
    mutate(text = paste(start, text, stop))
}

add_unknowns <- function(data, symbol = "_s_unk") {
  data %>%
    unnest_tokens(word, text) %>%
    group_by(word) %>%
    mutate(word2 = lag(word, n = 1, default = symbol)) %>%
    group_by(line) %>%
    summarize(text = paste(word2, collapse = " "))
}



count_trigrams <- function(data) {
  data %>%
    unnest_tokens(triple, text, token = "ngrams", n = 3, collapse = FALSE) %>%
    separate(triple, into = c("w1", "w2", "w3"), sep = " ", remove = FALSE) %>%
    count(w1, w2, w3) %>%
    ungroup()
}

build_unigram_model <- function(word_counts, k = 0.01) {
  word_counts %>%
    mutate(prob = (n + k) / (sum(n) + (k * n())))
}

build_bigram_model <- function(word_counts, bigram_counts, k = 0.01) {
  V <- nrow(word_counts)
  bigram_counts %>%
    inner_join(word_counts, by = c("w1")) %>%
    rename(c1 = n.y, c2 = n.x) %>%
    mutate(prob = (c2 + k) / (c1 + k * V)) %>%
    ungroup()
}

build_trigram_model <- function(bigram_counts, trigram_counts, k = 0.01) {
  V <- nrow(bigram_counts)
  trigram_counts %>%
    inner_join(bigram_counts, by = c("w1", "w2")) %>%
    rename(c1 = n.y, c2 = n.x) %>%
    mutate(prob = (c2 + k) / (c1 + k * V)) %>%
    ungroup()
}
```

## Data

Generate train-test-validate split.

```{r}
set.seed(42)

en_us_news_resample <- en_us_news %>%
  resample_partition(c(train = 0.80, validate = 0.10, test = 0.10))

en_us <- rbind(
  en_us_news,
  en_us_twitter,
  en_us_blogs
) %>%
  resample_partition(c(train = 0.80, validate = 0.10, test = 0.10))
```

Generate toy dataset.

```{r}
toy <- tribble(
  ~line, ~text,
  1, "I am Sam",
  2, "Sam I am",
  3, "I do not like green eggs and ham"
) %>%
  add_boundaries()
```

Set up train-test table.

```{r}
frac <- 1.0

news_train <- en_us_news_resample$train$data %>%
  sample_frac(frac) %>%
  add_boundaries() %>%
  add_unknowns()

news_validate <- en_us_news_resample$validate$data %>%
  sample_frac(frac) %>%
  add_boundaries()

us_train <- en_us$train$data %>%
  sample_frac(frac) %>%
  add_boundaries() %>%
  add_unknowns()

us_validate <- en_us$validate$data %>%
  sample_frac(frac) %>%
  add_boundaries()

tbl <- tribble(
  ~desc, ~train, ~validate,
  #"toy", toy, toy,
  #"news", news_train, news_validate,
  "us", us_train, us_validate
) %>%
  mutate(word_counts = map(train, count_words),
         unigram_model = map(word_counts, build_unigram_model),
         bigram_counts = map(train, count_bigrams),
         bigram_model = map2(word_counts, bigram_counts, build_bigram_model),
         trigram_counts = map(train, count_trigrams),
         trigram_model = map2(bigram_counts, trigram_counts, build_trigram_model))
tbl
```

Look at unigram model for toy set.

```{r}
tbl %>%
  filter(desc == "toy") %>%
  unnest(unigram_model) %>%
  arrange(desc(prob))
```

Look at bigram model for toy set.

```{r}
toy_bigrams <- tbl %>%
  filter(desc == "toy") %>%
  unnest(bigram_model)

toy_bigrams %>%
  arrange(desc(prob)) %>%
  head(10)

p_bigram <- function(y, x, data) {
  y <- tolower(y)
  x <- tolower(x)
  
  data %>%
    filter(w1 == x, w2 == y)
}
p_bigram("I", "_s_start", toy_bigrams) 
p_bigram("Sam", "_s_start", toy_bigrams)
p_bigram("am", "I", toy_bigrams)
p_bigram("_s_end", "Sam", toy_bigrams)
p_bigram("Sam", "am", toy_bigrams)
p_bigram("do", "I", toy_bigrams)
```

Look at unigram model.

```{r}
news_unigrams <- tbl %>%
  filter(desc == "news") %>%
  unnest(unigram_model)

news_unigrams %>%
  arrange(desc(prob)) %>%
  head(10)
```

Look at bigram model.

```{r}
news_bigrams <- tbl %>%
  filter(desc == "news") %>%
  unnest(bigram_model)

news_bigrams %>%
  arrange(desc(prob)) %>%
  head(10)
```

Look at trigram model.

```{r}
news_trigrams <- tbl %>%
  filter(desc == "news") %>%
  unnest(trigram_model)

news_trigrams %>%
  arrange(desc(prob)) %>%
  head(10)
```

Compute first-order Markov probability.
Studying computing of probability of "Sam I am".

```{r}
# P("Sam" | <s>)
p_sam_s <- news_bigrams %>%
  filter(w1 == "_s_start" & (w2 == "sam" | w2 == "_s_unk")) %>%
  arrange(desc(prob)) %>%
  .[["prob"]]
```

```{r}
# P("I" | "Sam")
m_i_sam <- news_bigrams %>%
  filter((w1 == "sam" | w1 == "_s_unk") & (w2 == "i" | w2 == "_s_unk")) %>%
  arrange(desc(prob))
m_i_sam
p_i_sam <- m_i_sam[[1, "prob"]]
```

**Q**: How do we choose between a partial match and a wholy unknown match?

```{r}
# P("am" | "I")
matches <- news_bigrams %>%
  filter((w1 == "i" | w1 == "_s_unk") & (w2 == "am" | w2 == "_s_unk")) %>%
  arrange(desc(prob))
matches
p_am_i <- matches[[1, "prob"]]
```

**Q**: How do we choose between partial and unknown matches and an explicit match with lower probability?

```{r}
# P(</s> | "am")
matches <- news_bigrams %>%
  filter((w1 == "am" | w1 == "_s_unk") & (w2 == "_s_end")) %>%
  arrange(desc(prob))
matches
p_s_am <- matches[[1, "prob"]]
```


```{r}
# P("Sam I am") = P("Sam" | <s>) * P("I" | "Sam") * P("am" | "I") * P(</s> | "am")
p_sam_s * p_i_sam * p_am_i * p_s_am
```

Implementing filter with a join.

```{r}
s <- tribble(
  ~ bigram, ~ w1, ~ w2,
  "<s> Sam", "_s_start", "sam",
  "<s> Sam", "_s_start", "_s_unk", # fill
  "Sam I", "sam", "i",
  "Sam I", "_s_unk", "i",          # fill
  "Sam I", "sam", "_s_unk",        # fill
  "Sam I", "_s_unk", "_s_unk",     # fill
  "I am", "i", "am",
  "I am", "_s_unk", "am",          # fill
  "I am", "i", "_s_unk",           # fill
  "I am", "_s_unk", "_s_unk",      # fill
  "am </s>", "am", "_s_end",
  "am </s>", "_s_unk", "_s_end"    # fill
)
s
s %>%
  left_join(news_bigrams, by = c("w1", "w2")) %>%
  group_by(bigram) %>%
  summarize(prob = max(prob, na.rm = TRUE)) %>%
  ungroup() %>%
  summarize(prob = prod(prob))
```

```{r}
tribble(
  ~ text,
  "Sam I am"
) %>%
  # Create bigrams
  add_boundaries() %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>%
  separate(bigram, into = c("w1", "w2"), sep = " ", remove = FALSE) %>%
  # Create all combinations
  mutate(u1 = ifelse(w1 != "_s_start", "_s_unk", NA)) %>%
  gather(k1, w1, w1, u1) %>%
  select(-k1) %>%
  na.omit() %>%
  select(bigram, w1, w2) %>%
  mutate(u2 = ifelse(w2 != "_s_end", "_s_unk", NA)) %>%
  gather(k2, w2, w2, u2) %>%
  select(-k2) %>%
  na.omit() %>%
  select(bigram, w1, w2) %>%
  arrange(bigram) %>%
  # Join with MLEs for bigrams
  left_join(news_bigrams, by = c("w1", "w2")) %>%
  # Filter out missing values
  na.omit() %>%
  # Pick the most-specific entry
  group_by(bigram) %>%
  arrange(desc(w1), desc(w2), desc(prob)) %>%
  mutate(row = row_number()) %>%
  filter(row == 1) %>%
  # Pick the entries with highest probability
  #group_by(bigram) %>%
  #arrange(desc(prob)) %>%
  #mutate(row = row_number()) %>%
  #filter(row == 1) %>%
  # Compute probability
  summarize(prob = max(prob)) %>%
  ungroup() %>%
  summarize(prob = prod(prob))
```

Calculating perplexit using validation set.

```{r}
# M = total #(words) in hold-out
M <- news_validate %>%
  unnest_tokens(word, text) %>%
  count(word) %>%
  summarize(total = sum(n)) %>%
  .[["total"]]
M

# perplexit(s_i)
s1 <- news_validate %>%
  select(text) %>%
  head(1) %>%
  .[["text"]]
s1

s2 <- news_train %>%
  select(text) %>%
  sample_n(1) %>%
  .[["text"]]
s2
```

```{r}
tribble(
  ~ text,
  "_s_start while happy owners were helping customers the little dog who looks to be a jerk jumped onto a chair and then onto a desk on top of the desk were four packs of dog treats _s_end"
) %>%
  # Create bigrams
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>%
  separate(bigram, into = c("w1", "w2"), sep = " ", remove = FALSE) %>%
  # Create all combinations
  mutate(u1 = ifelse(w1 != "_s_start", "_s_unk", NA)) %>%
  gather(k1, w1, w1, u1) %>%
  select(-k1) %>%
  na.omit() %>%
  select(bigram, w1, w2) %>%
  mutate(u2 = ifelse(w2 != "_s_end", "_s_unk", NA)) %>%
  gather(k2, w2, w2, u2) %>%
  select(-k2) %>%
  na.omit() %>%
  select(bigram, w1, w2) %>%
  arrange(bigram) %>%
  # Join with MLEs for bigrams
  left_join(news_bigrams, by = c("w1", "w2")) %>%
  # Filter out missing values
  na.omit() %>%
  # Pick the most-specific entry
  group_by(bigram) %>%
  arrange(desc(w1), desc(w2), desc(prob)) %>%
  mutate(row = row_number()) %>%
  filter(row == 1) %>%
  # Pick the entries with highest probability
  #group_by(bigram) %>%
  #arrange(desc(prob)) %>%
  #mutate(row = row_number()) %>%
  #filter(row == 1) %>%
  # Compute probability
  summarize(prob = max(prob)) %>%
  ungroup() %>%
  summarize(prob = prod(prob)) %>%
  .[["prob"]]
```

Assess quality of unigram model.

```{r}
# Calculate perplexity.
#
# Assume that test set has start and end markers.
perplexity <- function(test, model) {
  test_words <- test %>%
    unnest_tokens(word, text)
    
  total_count <- test_words %>%
    filter(word != "_s_start") %>%
    distinct(word) %>%
    count() %>%
    .[["n"]]

  pp <- test_words %>%
    inner_join(model, by = c("word" = "w1")) %>%
    mutate(log_prob = log(prob)) %>%
    summarize(sum_log_prob = sum(log_prob)) %>%
    mutate(prod_prob = exp(sum_log_prob),
           pp = prod_prob^(-1 / total_count))
  
  pp
}

news_validate %>%
  unnest_tokens(word, text)

pp_test <- perplexity(news_validate, news_unigrams)
pp_test
pp_test <- perplexity(toy, news_unigrams)
```

## Summary

Bigram model with add-k smoothing.

## Prototype

Given a tidy dataset of lines of text, produce bigram probabilities.

* Add unknown words
    * "replace the first occurrence of every word type in the training data by <UNK>" (Jurafsky)

```{r}
```



```{r}

# Predict follow-on from single word
w <- "an"
ex1_w <- ex1_pk %>%
  filter(left == w) %>%
  filter(!(right %in% c("s1", "s2", "s3"))) %>%
  arrange(desc(pk))
ex1_w %>%
  sample_n(3, weight = pk)

# Predict follow-on from sentence
s <- "I want an"
w <- str_trim(str_extract(s, "\\s\\w+$"))
ex1_s <- ex1_pk %>%
  filter(left == w) %>%
  filter(!(right %in% c("s1", "s2", "s3"))) %>%
  arrange(desc(pk))
ex1_w %>%
  sample_n(3, weight = pk) %>%
  transmute(s = paste(s, right))
```

### Trigram model

Using trigram model for Quiz 2.

```{r}
us_trigrams <- tbl %>%
  filter(desc == "us") %>%
  unnest(trigram_model)

quiz_2 <- tribble(
  ~ question, ~ text, ~ a, ~ b, ~ c, ~ d,
  1, "When you breathe, I want to be the air for you. I'll be there for you, I'd live and I'd", "sleep", "die", "give", "eat",
  2, "Guy at my table's wife got up to go to the bathroom and I asked about dessert and he started telling me about his", "marital", "financial", "horticultural", "spiritual",
  3, "I'd give anything to see arctic monkeys this", "weekend", "month", "morning", "decade",
  4, "Talking to your mom has the same effect as a hug and helps reduce your", "sleepiness", "stress", "hunger", "happiness",
  5, "When you were in Holland you were like 1 inch away from me but you hadn't time to take a", "walk", "picture", "minute", "look",
  6, "I'd just like all of these questions answered, a presentation of evidence, and a jury to settle the", "incident", "account", "matter", "case",
  7, "I can't deal with unsymetrical things. I can't even hold an uneven number of bags of groceries in each", "arm", "finger", "toe", "hand",
  8, "Every inch of you is perfect from the bottom to the", "center", "top", "middle", "side",
  9, "I’m thankful my childhood was filled with imagination and bruises from playing", "inside", "outside", "weekly", "daily",
  10, "I like how the same people are in almost all of Adam Sandler's", "stories", "novels", "pictures", "movies"
) %>%
  gather(option, value, a:d) %>%
  unite(text, text, value, sep = " ") %>%
  unnest_tokens(triples, text, token = "ngrams", n = 3) %>%
  group_by(question, option) %>%
  filter(n() - row_number() == 0) %>%
  separate(triples, into = c("w1", "w2", "w3"), sep = " ") %>%
  mutate(u1 = "_s_unk") %>%
  gather(k1, w1, w1, u1) %>%
  select(-k1) %>%
  mutate(u2 = "_s_unk") %>%
  gather(k2, w2, w2, u2) %>%
  select(-k2) %>%
  ungroup() %>%
  left_join(us_trigrams, by = c("w1", "w2", "w3")) %>%
  na.omit() %>%
  select(question, option, w1, w2, w3, prob) %>%
  arrange(question, desc(prob))
quiz_2
```

```{r}
# Count trigrams
ex1_c3 <- ex1 %>%
  unnest_tokens(pair, text, token = "ngrams", n = 3, collapse = FALSE) %>%
  separate(pair, into = c("w1", "w2", "w3"), sep = " ", remove = FALSE) %>%
  count(w1, w2, w3) %>%
  rename(c3 = n)
ex1_c3 %>%
  arrange(desc(c3))

# Compute add-k smoothing
k <- 0.01
V <- nrow(ex1_c2)
ex1_pk <- ex1_c3 %>%
  inner_join(ex1_c2, by = c("w1" = "left", "w2" = "right")) %>%
  mutate(pk = (c3 + k) / (c2 + k * V)) %>%
  ungroup()
ex1_pk %>%
  arrange(desc(pk))

# Predict follow-on from sentence
predict_bigram <- function(data, str, start = "s1", stop = "s2", symbol = "s3") {
  words <- rev(rev(str_split(str, " ")[[1]])[1:2])
  bigram_data <- data %>%
    filter(w1 %in% c(ws[[1]], symbol)) %>%
    filter(!(w2 %in% c("s1", "s2", "s3")))
}

# Predict follow-on from sentence
predict_trigram <- function(data, strs, start = "s1", stop = "s2", symbol = "s3") {
  text <- strs %>%
    unnest_tokens(word, text) %>%
    mutate(w = paste0('t', row_number() - n() + 3)) %>%
    spread(w, word)
  
  text %>%
    left_join(data, by = c("t1" = "w1", "t2" = "w2"))
}
```

## Evaluation


Build maximum likelihood estimation (MLE) based on sample data.

```{r}
bigram_probs <- en_us_news_resample$train$data %>%
  sample_frac(0.10) %>%
  ym_1c()

str(bigram_probs)
bigram_probs
object_size(bigram_probs)
summary(bigram_probs)
```

```{r}
bigram_probs %>%
  filter(left == "than") %>%
  arrange(desc(mle))
```

```{r}
bigram_probs %>%
  filter(right == "321e")
```

### Toy sample

```{r}
p_sentence <- function(data, s) {
  s <- paste("s123", s, "321e")
  df <- data.frame(text = s, stringsAsFactors = FALSE) %>%
    unnest_tokens(pair, text, token = "ngrams", n = 2, collapse = FALSE) %>%
    separate(pair, into = c("left", "right"), sep = " ") %>%
    left_join(data, by = c("left", "right")) %>%
    summarize(prob = prod(mle))
  
  df
}

p_sentence(toy_probs, "Sam I am")
p_sentence(toy_probs, "Hi there")
```

